This dissertation has been a thorough investigation of the advantages and
disadvantages of a shared-environment approach to implementing call-by-need
semantics. In Chapter~\ref{chap:cem}, I investigated the runtime efficiency
advantages by implementing a simple native code compiler. Despite the compilers
lack of optimization framework, it was often competitive with the state of the
art. From this, I concluded that this approach is a promising abstract machine 
for real-world compilers. In Chapter~\ref{chap:verified}, I showed how I could
use the simplicity of the implementation to effectively reason formally about
its correctness. While it was a significant undertaking, the success of the
verified compiler provides strong evidence that the simplicity of the machine is
a valuable property, and that property can be further exploited in future work.

For the rest of this Chapter, I am retrospective on the work done for the
dissertation, discussing both worked well and what didn't. The hope is that this
deeper dive into the challenges and successes throughout the dissertation can
better inform future work, both work that uses the \ce machine directly, and
work that might take a different approach, but with similar goals. While some of
what is discussed in this section has been covered in Chapter-specific
conclusions from previous Chapters, we try and unify the conclusions into
bigger-picture conclusions, accumulating the lessons learned along the way. In
addition, through discussions with other experts and through further
introspection, more conclusions and lessons learned have been discovered, and we
use this section to bring those to light. 

This chapter is separated into a few important subsections. First, it summarizes
the theses and conclusions of the dissertation. Second, it summarizes and
addresses the threats to validity of the conclusions presented. Largely
motivated by addressing these threats to validity, it then discusses future
directions enabled by this work. 

\section{Theses and Their Conclusions}

There are two core theses presented in this dissertation. First,
shared-environments efficiently implement call-by-need semantics, as described
by the \ce machine. This thesis is described in detail in
Chapter~\ref{chap:cem}. The conclusion of this chapter is that there are clear
efficiencies gained and efficiencies lost with the shared environment approach.
In particular, the efficiency gained by reduced thunk creation overhead enabled
techniques like more efficient scott-encoded datatypes. 

The primary conclusion for this thesis is that this is a good default behavior,
but one that should be optimized away. Consider the analogue of strictness
analysis. The philosophy is that laziness is a good default, but one that should
be removed for efficiency where possible. In the same way, lazy thunk creation
is a good default, but one that should be replaced with an optimized flat
environment closure where necessary for efficiency reasons. In terms of cheap to
create vs. efficient to execute, we can think of the shared-environment closure
as the most extremely cheap to create, at the cost of efficiency to execute,
where as a just-in-time compiled closure specialized on its values might be at
the other extreme: expensive to create, but efficient to execute. 

The second thesis of the dissertation is that the simplicity of the compiler
that implements the \ce machine lends itself to formal reasoning. This thesis is
described in detail in Chapter~\ref{chap:verified}. The evidence for this thesis
is provided in the form of a verified compiler. By implementing the first
verified compiler of a call-by-need semantics, I have provided compelling
support for this thesis. 

The conclusion for this thesis is therefore that yes, the simplicity of the
compiler indeed lends itself to formal reasoning. That said, there are still
many open questions. For example, many of the proofs are excruciatingly complex.
It is therefore not at all clear that the structure of the compiler and proofs
is \emph{the best possible}. Indeed, we expect there is a lot of room for
improvement, particularly in the implementation of the proofs. We discuss these
possibilities further in the next section. 

\section{Threats to Validity}

This section attempts to summarize and address what it considers to be the most
pressing threats to the validity to the theses and conclusions discussed in this
dissertation. While it attempts to enumerate the most pressing threats, any list
of this nature will be incomplete, and therefore the goal is not to create a
complete list. Instead, the aim of the section is simply to convey that possible
criticisms have been seriously considered. Note that Chapters~\ref{chap:cem} and
\ref{chap:verified} have chapter-specific threats to validity, so in this
section we instead focus on more general threats to validity. 

The first, and most glaring, is that the verified compiler makes no claims about
the behaviour of the compiled code in the case of non-termination in the source
semantics. This was discussed thoroughly in Chapter~\ref{chap:verified}, but
it is worth returning to here. It is a hole that I am comfortable, due largely
to the belief that effectively everyone will be programming in total languages
in the future, making this hole disappear. This belief was not represented in
the dissertation in any way, but even if we accept that this work only applies
to total languages, we can accept that they do exist, and in the context of
those languages, this is a complete verified compiler. 

Another sort of existential threat to validity is that I haven't built a
\emph{true} real world compiler yet, and therefore can make no claims about the
performance or simplicity of such a beast. Part of this comes from a vague
definition of what it means to be a real world compiler. To paraphrase Edwin
Brady, the creator of the Idris programming language, a language should be
"pacman-complete", that is, it should be straightforward to implement pacman in
the language. In this way, a compiler should be able to compile such a program.
While I would claim that the native code compiler could indeed compile such a
program, I also accept that a much better solution would be to have it compile
full Haskell, as well as have a fully functional, high-performance garbage
collector integrated into the generated code. While pacman-completeness is more
about source language, there is also disagreement about what constitutes a
real-world compiler in terms of what the internal languages implement. For
example, many are inclined to dismiss a compiler, like the one presented here,
that doesn't implement recursion and algebraic data types in the internal
language. A tangential goal of this dissertation is to open the reader to the
possibility that those are not necessary for high performance code. Removing
them certainly makes formal reasoning simpler, something that hopefully the
reader is now convinced is an important property for a compiler to have.  

\section{Future Work}

While Chapters~\ref{chap:cem} and \ref{chap:verified} have sections dedicated to
future work specific to their topics, I expand on those here, discussing future
work that could combine the approaches of both efforts. 

The most obvious one would be to combine the efforts. By extracting the Coq
implementation of the compiler into Haskell, the language that the native code
compiler is implemented in, we would attain a verified fragment of a true native
code compiler. We could then work towards extending the proof to cover more of
the implementation over time. When combined with the possibility of implementing
a full Haskell compiler using the native code compiler in this work, we have a
viable path towards a high performance, fully verified compiler for Haskell. 

This approach neglects the fact that there are many cases where GHC
significantly outperforms the native code compiler in its current form. Recent
work on verified transformations and optimizations, when combined with our
verified compiler, could result in a verified compiler that is actually
competitive with GHC in performance, likely outperforming it in some cases due
to lightweight closure creation enabled by the \ce machine. In a sense, this
would be a natural next step to the existing type-safe core language of GHC:
instead of just ensuring that transformations are type-safe, we could have a
compiler that ensures that they are correct.

While discussed briefly in the previous chapters, one exciting area for future
work is reasoning formally about performance. Reasoning about memory use of lazy
functional programs is notoriously hard, so any help that the compiler can
provide is extremely valuable. Unfortunately, heuristic approaches interacting
with the many optimizations that exist already can lead to extreme difficulty in
reasoning about memory consumption of a source program. We hope that in
combination with future language tools, one could use the simplicity of the
compiler presented here to better reason about time and space efficiencies, and
help ensure that the compiler makes better decisions about how to preserve
and/or improve time and space requirements. The lack of many intermediate
representations make this compiler an ideal target for such reasoning.

\section{Conclusion}

I have presented a pair of compilers that provide evidence for the two main
claims of this dissertation: the \ce abstract machine gives rise to efficient,
easy to reason about compilers. 

If the reader accepts the claims that those are desirable properties for a
compiler to have, then she must conclude that the \ce machine and the compiler
it enables are valuable tools for implementing call-by-need for real-world
projects. The evidence accumulated throughout the dissertation compels me to
to agree with the reader that yes, the work presented here can be used in
real-world implementations. Both properties discussed, efficiency of closure
creation and ability to reason formally about compilers, are clearly valuable
properties to have for a compiler. Therefore, we can conclude that the work
presented in this dissertation will prove useful when attempting to implement a
lazy compiler with those properties. 


