\documentclass[botnum]{unmeethesis}
\usepackage{mathtools}
\usepackage{tabularx}
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tikz}
\usetikzlibrary{chains,fit,shapes,calc}
\usepackage{verbatim}
\usepackage{semantic}
\usepackage{tabu}
\usepackage{mathptmx}
\usepackage{todonotes}
\usepackage{amsthm}
\usepackage{amssymb}
\usepackage{hyperref}
\usepackage{epigraph}
\usepackage{fontspec}
\usepackage{amsmath}
\usepackage{rotating}
\usepackage{caption}

\setmonofont[Scale=0.8]{DejaVu Sans Mono}

\newcommand{\concat}{\ensuremath{+\!\!\!\!+\,}}  
\newtheorem{thm}{Theorem}
\newtheorem{cor}{Corollary}
\newtheorem{lem}{Lemma}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newenvironment{proofoutline}
 {\renewcommand\qedsymbol{}\proof[Proof outline]}
 {\endproof}
\def\ce{$\mathcal{\mathcal{C} \mskip -4mu \mathcal{E}} \mskip 4mu$}

\begin{document}

\frontmatter

% Uncomment the next command if you see weird paragraph spacing:
% That is, if you see paragraphs float with lots of white space
% in between them:

% \setlength{\parskip}{0.30cm}

\title{Shared-Environment Call-by-Need}

\author{George Widgery Stelle}

\degreesubject{Ph.D., Computer Science}

\degree{Doctor of Philosophy \\ Computer Science}

\documenttype{Dissertation}

\previousdegrees{B.S., University of British Columbia, 2008 \\
                 M.S., Computer Science, University of New Mexico, 2013}

\date{April, \thisyear}

\maketitle

%\makecopyright

\begin{dedication}
For Beth 
\end{dedication}

\begin{acknowledgments}
  \vspace{1.1in}
  The list of people who deserve acknowledgment for this dissertation is far too
  long to enumerate here. Instead, I'll make a meager attempt at mentioning a
  few who played particularly important roles. 

  Starting at the beginning, I'd like to thank my parents and brothers, who gave
  me a childhood rich in love and fun. In a sense, my decision to pursue a Ph.D.
  was driven by a need to continue having fun doing what I love.  

  Most of the fun I've had has not been sitting and thinking about hard
  problems, but making lifelong friends along the way. To Taylor, Drew, Ben,
  Eric, George, Vu, and others, thank you for the technical discussions, the
  parties, the games, the hikes, the road trips, and most of all your
  friendship. Also, thanks to my three amigos for life, Drake, Mike, and Larkin,
  for the much-needed breaks from work. 

  I thank my Ph.D. advisor, Darko Stefanovic, for his support and advice through
  my unconventional final years through the program. Without his help and
  wisdom there is no question that this dissertation would not exist. Stephanie
  Forrest deserves a great deal of credit as well, both for bringing me into the
  department, and supporting and advising me selflessly through the challenging
  task of finding a topic I love. I would not be here in New Mexico at all
  without the help of my friend Eric Vatikiotis-Bateson. 

  The papers that form this basis of this dissertation would not have been
  possible without the help of my extraordinary co-authors, Darko Stefanovic,
  Stephen Olivier, and Stephanie Forrest. Their ability to write clearly and
  carefully continues to inspire me to be a better communicator. My committee
  members, which include the above co-authors, as well as Kei Davis and Patrick
  Bridges, have my thanks for reviewing my dissertation and sitting through my
  defense. 

  I thank my wonderful wife, Beth, for being my partner in life and always
  inspiring me to be better. And to my children, Adelaide and Sullivan, thank
  you for making life so much more.
\end{acknowledgments}

\maketitleabstract %(required even though there's no abstract title anymore)

\begin{abstract}
Call-by-need semantics formalizes the wisdom that work should be done at most
once. It frees programmers to focus more on the correctness of their code, and
less on the operational details. As a result, programmers of lazy functional
languages rely particularly heavily on their compiler to preserve correctness,
while also relying on the compiler to generate high performance code for high
level abstractions. In this dissertation, I present a novel technique for
compiling call-by-need semantics that uses shared environments to share
results of computation. I show how the approach enables a compiler that
generates high performance code, while staying simple enough to lend itself to
formal reasoning. The dissertation is divided into three main contributions.
First, I present an abstract machine, the \ce machine, which formalizes the
approach.  Second, I show that it can be implemented as a native code compiler
with encouraging performance results.  Finally, I present a verified compiler,
implemented in the Coq proof assistant, demonstrating how the simplicity of
the approach enables formal verification.
\clearpage %(required for 1-pageabstract)
\end{abstract}

\tableofcontents
\listoffigures
%\listoftables

\mainmatter

\chapter{Introduction}\label{chap:intro}
\epigraph{To be admitted to Nature's hearth costs nothing. None is excluded, but
excludes himself. You have only to push aside the curtain.}{\textit{Henry David
Thoreau}}
\input{intro}

\chapter{Background}\label{chap:background}
\epigraph{Science is the belief in the ignorance of experts.}{\textit{Richard
Feynman}}
\input{background}

\chapter{\ce Machine}\label{chap:ce}
\epigraph{The lurking suspicion that something could be simplified is the
world's richest source of rewarding challenges.}{\textit{Edsger Dijkstra}}
\input{ce}
\input{tfp16/calc}
\input{tfp16/mach}

\chapter{Native Code Compilation}\label{chap:cem}
\epigraph{Much of my work has come from being lazy.}{\textit{John Backus}}
\input{tfp16/intro}
\input{tfp16/implement}
\input{tfp16/evaluation}
\input{tfp16/results}
\input{tfp16/discussion}

\chapter{Verified Compilation}\label{chap:verified}
\epigraph{I don't believe in empirical science. I only believe in a priori
truth.}{\textit{Kurt Gödel}}

\input{ifl18/intro}
\input{ifl18/cem}
\input{ifl18/cesm}
\input{ifl18/im}
\input{ifl18/compiler}
\input{ifl18/correctness}
\input{ifl18/discussion}

\chapter{Conclusions}\label{chap:conclusion}
\epigraph{Understand as well as I may, my comprehension can only be an
infinitesimal fraction of all I want to understand.}{\textit{Ada Lovelace}}
%\epigraph{He who learns but does not think, is lost! He who thinks but does not
%learn is in great danger.}{\textit{Confucius}}
\input{conclusions}

\chapter*{Appendices}

\appendix
\chapter{Native Code Compiler Implementation Details}\label{chap:cem_appendix}
\input{cem_appendix} 

\chapter{Coq Implementation Details}\label{chap:coq_appendix}
\input{coq_appendix} 

\bibliographystyle{amsplain}
\bibliography{annotated}

\end{document}
